---
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Probability and Statistics

# Lab Assignment 1: Naive Bayes Classifier

## Work breakdown

-   *Name1 Surname1*: Oleksandr Ivaniuk
-   *Name2 Surname2*: Olena Hnyp
-   *Name3 Surname3*: Dmytro Hamula

## Introduction

During the first three weeks, you learned a couple of essential notions
and theorems, and one of the most important among them is the *Bayes
theorem*.

One of its applications is **Naive Bayes classifier**, which is a
probabilistic classifier whose aim is to determine which class some
observation probably belongs to by using the Bayes formula:
$$\mathsf{P}(\mathrm{class}\mid \mathrm{observation})=\frac{\mathsf{P}(\mathrm{observation}\mid\mathrm{class})\mathsf{P}(\mathrm{class})}{\mathsf{P}(\mathrm{observation})}$$

Under the strong independence assumption, one can calculate
$\mathsf{P}(\mathrm{observation} \mid \mathrm{class})$ as
$$\mathsf{P}(\mathrm{observation}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i), \qquad \mathsf{P}(\mathrm{observation} \mid \mathrm{class}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i \mid \mathrm{class}),$$
where $n$ is the total number of features describing a given
observation. Thus, $\mathsf{P}(\mathrm{class}|\mathrm{observation})$ now
can be calculated as

$$\mathsf{P}(\mathrm{class} \mid \mathrm{\mathrm{observation}}) = \mathsf{P}(\mathrm{class})\times \prod_{i=1}^{n}\frac{\mathsf{P}(\mathrm{feature}_i\mid \mathrm{class})}{\mathsf{P}(\mathrm{feature}_i)}\tag{1}$$

All the terms on the right-hand side can be estimated from the data as
respective relative frequencies;\
**\*see [this
site](https://monkeylearn.com/blog/practical-explanation-naive-bayes-classifier/)
for more detailed explanations\***

## Data description

There are 5 datasets uploaded on the cms (data.zip)

To determine your variant, take your team number from the list of teams
on cms and take *mod 5* - this is the number of your data set.

-   **0 - authors** This data set consists of citations of three famous
    writers: Edgar Alan Poe, Mary Wollstonecraft Shelley and HP
    Lovecraft. The task with this data set is to classify a piece of
    text with the author who was more likely to write it.

-   **1 - discrimination** This data set consists of tweets that have
    discriminatory (sexism or racism) messages or of tweets that are of
    neutral mood. The task is to determine whether a given tweet has
    discriminatory mood or does not.

-   **2 - fake news** This data set contains data of American news: a
    headline and an abstract of the article. Each piece of news is
    classified as fake or credible. The task is to classify the news
    from test.csv as credible or fake.

-   **3 - sentiment** All the text messages contained in this data set
    are labeled with three sentiments: positive, neutral or negative.
    The task is to classify some text message as the one of positive
    mood, negative or neutral.

-   **4 - spam** This last data set contains SMS messages classified as
    spam or non-spam (ham in the data set). The task is to determine
    whether a given message is spam or non-spam.

Each data set consists of two files: *train.csv* and *test.csv*. The
first one you will need find the probabilities distributions for each of
the features, while the second one is needed for checking how well your
classifier works.

```{r}
# here goes a list of recommended libraries,
# though you may install other ones if they are needed
library(tidytext)
library(readr)
library(dplyr)
library(ggplot2)
library(ggwordcloud)
```

## Outline of the work

1.  **Data pre-processing** (includes removing punctuation marks and
    stop words, representing each message as a bag-of-words)
2.  **Data visualization** (it's time to plot your data!)
3.  **Classifier implementation** (using the training set, calculate all
    the conditional probabilities in formula (1) and then use those to
    predict classes for messages in the testing set)
4.  **Measurements of effectiveness of your classifier** (accuracy,
    precision and recall curves, F1 score metric etc)
5.  **Conclusions**

*!! do not forget to submit both the (compiled) Rmd source file and the
.html output !!*

## Data pre-processing

-   Read the *.csv* data files.
-   Сlear your data from punctuation or other unneeded symbols.
-   Clear you data from stop words. You don't want words as is, and, or
    etc. to affect your probabilities distributions, so it is a wise
    decision to get rid of them. Find list of stop words in the cms
    under the lab task.
-   Represent each test message as its bag-of-words. Here:
    <https://machinelearningmastery.com/gentle-introduction-bag-words-model/>
    you can find general introduction to the bag-of-words model and
    examples on to create it.
-   It is highly recommended to get familiar with R dataframes, it would
    make the work much easier to do.
-   Useful links:
    -   <https://steviep42.github.io/webscraping/book/bagofwords.html#tidytext> -
        example of using *tidytext* to count frequencies of the words.
    -   Basics of Text Mining in R:
        <http://rstudio-pubs-static.s3.amazonaws.com/256588_57b585da6c054349825cba46685d8464.html>
        . Note that it also includes an example on how to create a bag
        of words from your text document.

```{r}
list.files(getwd())
list.files("data/4-spam")
```

```{r}
test_path <- "data/4-spam/test.csv"
train_path <- "data/4-spam/train.csv"

stop_words <- read_file("stop_words.txt")
# https://stackoverflow.com/questions/27195912/why-does-strsplit-return-a-list
splitted_stop_words <- strsplit(stop_words, split='\n')
splitted_stop_words <- splitted_stop_words[[1]]

```

```{r}
train <-  read.csv(file = train_path, stringsAsFactors = FALSE)
test <-  read.csv(file = test_path, stringsAsFactors = FALSE)
print(train)
```

```{r}
# note the power functional features of R bring us! 

tidy_text_training <- train %>%
                      mutate(Category = ifelse(Category == "spam", 1, 0)) %>%
                      unnest_tokens('splitted', 'Message', token = "lines")
tidy_text_test <- test %>%
                      mutate(Category = ifelse(Category == "spam", 1, 0)) %>%
                      unnest_tokens('splitted', 'Message', token = "lines")

training_X = tidy_text_training$splitted
training_y = tidy_text_training$Category

test_X = tidy_text_test$splitted
test_y = tidy_text_test$Category

tidy_text <- unnest_tokens(train, 'splitted', 'Message', token="words") %>%
             filter(!splitted %in% splitted_stop_words)

spam_df <- tidy_text[tidy_text$Category == "spam", ]
word_counts_spam <- spam_df %>% count(splitted, sort = TRUE)
spam_df <- left_join(spam_df, word_counts_spam, by = "splitted")
spam_df <- spam_df %>% distinct()
spam_df <- spam_df %>%
  arrange(desc(n))

ham_df <- tidy_text[tidy_text$Category == "ham", ]
word_counts_ham <- ham_df %>% count(splitted, sort = TRUE)
ham_df <- left_join(ham_df, word_counts_ham, by = "splitted")
ham_df <- ham_df %>% distinct()
ham_df <- ham_df %>%
  arrange(desc(n))

word_counts <- tidy_text %>% count(splitted, sort = TRUE)
tidy_text <- left_join(tidy_text, word_counts, by = "splitted")
tidy_text$Category <- "ham + spam"
tidy_text <- tidy_text %>% distinct()
tidy_text <- tidy_text %>%
  arrange(desc(n))
tidy_text

spam_df
ham_df
```

## Data visualization

Each time you work with some data, you need to understand it before you
start processing it. R has very powerful tools to make nice plots and
visualization. Show what are the most common words for negative and
positive examples as a histogram, word cloud etc. Be creative!

```{r}
first_30_rows_spam <- head(spam_df, 30)
set.seed(42)
ggplot(first_30_rows_spam, aes(label = splitted, size = n, color = factor(sample.int(10, nrow(first_30_rows_spam), replace = TRUE)))) +
  scale_size_area(max_size = 20) +
  geom_text_wordcloud() +
  theme_minimal()
```

```{r}
first_30_rows_ham <- head(ham_df, 30)
set.seed(42)
ggplot(first_30_rows_ham, aes(label = splitted, size = n, color = factor(sample.int(10, nrow(first_30_rows_ham), replace = TRUE)))) +
  scale_size_area(max_size = 20) +
  geom_text_wordcloud() +
  theme_minimal()
```

```{r}
ggplot(data = first_30_rows_spam, aes(x = splitted, y = n)) +
  geom_bar(stat = "identity", fill = "red") +
  xlab("Spam words") +
  ylab("Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(data = first_30_rows_ham, aes(x = splitted, y = n)) +
  geom_bar(stat = "identity", fill = "green") +
  xlab("Ham words") +
  ylab("Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Classifier implementation

```{r}
naiveBayes <- setRefClass("naiveBayes",
                          
       # here it would be wise to have some vars to store intermediate result
       # frequency dict etc. Though pay attention to bag of words!
       fields = list(incorrect = "data.frame", df_test = "data.frame"),
       methods = list(
                    # prepare your training data as X - bag of words for each of your
                    # messages and corresponding label for the message encoded as 0 or 1 
                    # (binary classification task)
                    fit = function(X, y)
                    {
                        df <- data.frame(
                            Category = y,
                            Message = X
                        )
                        df <- unnest_tokens(df, "splitted", "Message", token = "words") %>%
                                        filter(!(splitted %in% splitted_stop_words))
                        spam_df_train <- df[df$Category == "1", ]
                        word_counts_spam <- spam_df_train %>% count(splitted, sort = TRUE)
                        spam_df_train <- left_join(spam_df_train, word_counts_spam, by = "splitted")
                        spam_df_train <- spam_df_train %>% distinct()
                        spam_df_train <- spam_df_train %>%
                            arrange(desc(n))
                        
                        ham_df_train <- df[df$Category == "0", ]
                        word_counts_ham <- ham_df_train %>% count(splitted, sort = TRUE)
                        ham_df_train <- left_join(ham_df_train, word_counts_ham, by = "splitted")
                        ham_df_train <- ham_df_train %>% distinct()
                        ham_df_train <- ham_df_train %>%
                          arrange(desc(n))
                        
                        word_counts_train <- df %>% count(splitted, sort = TRUE)
                        df <- left_join(df, word_counts, by = "splitted")
                        tidy_text$Category <- "0 + 1"
                        df <- df %>% distinct()
                        df <- df %>%
                          arrange(desc(n))
                        
                        spam_dict <- list()
                        ham_dict <- list()
                        probabilities <- list()
                        neg_prob <- list()
                        cnt_words <- sum(word_counts_train$n)
                        
                        for (i in 1:nrow(ham_df_train)) {
                          key <- ham_df_train$splitted[i]
                          value <- ham_df_train$n[i]
                          ham_dict[[key]] <- value
                        }                        
                        for (i in 1:nrow(spam_df_train)) {
                          key <- spam_df_train$splitted[i]
                          value <- spam_df_train$n[i]
                          spam_dict[[key]] <- value
                        }
                        for (word in word_counts_train$splitted) {
                          if (exists(word, where = spam_dict)){
                            probabilities[[word]] <- (spam_dict[[word]])/(cnt_words)
                          }
                          else{
                            probabilities[[word]] <- 1/cnt_words
                          }
                          if (exists(word, where = ham_dict)){
                            neg_prob[[word]] <- (ham_dict[[word]])/(cnt_words)
                          }
                          else{
                            neg_prob[[word]] <- 1/cnt_words
                          }
                        }
                        ret_list <- list(
                          "first" = probabilities,
                          "second" = neg_prob
                        )
                        
                        return(ret_list)
                    },
                    
                    # return prediction for a single message 
                    predict = function(message, probabilities, neg_prob)
                    {
                        cleaned_message <- gsub("[[:punct:]]", " ", message)
                         words <- unlist(strsplit(cleaned_message, " "))
                         words <- subset(words, !words == "")
                         prob_0 <- 1
                         prob_1 <- 1
                         for (word in words){
                           if (word %in% names(probabilities)){
                              prob_1 <- prob_1 * probabilities[[word]]
                           }
                           else if (word %in% names(neg_prob)){
                             prob_1 <- prob_1 * neg_prob[[word]]
                           }
                           if (word %in% names(neg_prob)){
                              prob_0 <- prob_0 * neg_prob[[word]]
                           }
                           else if (word %in% names(probabilities)){
                             prob_0 <- prob_0 * neg_prob[[word]]
                           }
                         }
                        if (prob_0 > prob_1) {
                           return(0)
                         }
                         else {
                           return(1)
                         }
                    },
                    
                    # score you test set so to get the understanding how well you model
                    # works.
                    # look at f1 score or precision and recall
                    # visualize them 
                    # try how well your model generalizes to real world data! 
                    score = function(X_test, y_test)
                    {
                        model = naiveBayes()
                        p_list <- model$fit(training_X, training_y)
                        df_test <<- data.frame(
                          Category = y_test,
                          Message = X_test
                        )
                        df_test <<- unnest_tokens(df_test, "splitted", "Message", token = "lines")
                        incorrect <<- .self$incorrect 
                        messages_total <- nrow(df_test)
                        predicted_correct <- 0
                        
                        true_positives <- 0 #correctly predicted spam
                        true_negatives <- 0 #correctly predicted ham
                        false_positives <- 0 #predicted spam but was ham
                        false_negatives <- 0 #predicted ham but was spam
                        
                        for (i in 1:nrow(df_test)) {
                          df_test$predicted[i] <<- paste(predict(df_test$splitted[i], p_list[["first"]], p_list[["second"]]))
                        } 
                        for (j in 1:nrow(df_test)) {
                          if (as.integer(df_test$Category[j]) == as.integer(df_test$predicted[j])) {
                            predicted_correct <- predicted_correct + 1
                          }
                          else {
                            incorrect <<- rbind(incorrect, df_test$splitted[j])
                          }
                          
                          if (as.integer(df_test$Category[j]) == 1 & as.integer(df_test$predicted[j] == 1)) {
                            true_positives <- true_positives + 1
                          }
                          else if (as.integer(df_test$Category[j]) == 0 & as.integer(df_test$predicted[j] == 1)) {
                            false_positives <- false_positives + 1
                          }
                          else if (as.integer(df_test$Category[j]) == 1 & as.integer(df_test$predicted[j] == 0)) {
                            false_negatives <- false_negatives + 1
                          }
                          else if (as.integer(df_test$Category[j]) == 0 & as.integer(df_test$predicted[j] == 0)) {
                            true_negatives <- true_negatives + 1
                          }
                          
                        }
                        colnames(incorrect) <<- "Incorrect"
                        .self$df_test <- df_test
                        #calculate f1
                        p <- true_positives / (true_positives + false_positives) #precision
                        r <- true_positives / (true_positives + false_negatives) #recall
                        f1 <- (2 * p * r) / (p + r)
                        #calculate score
                        score <- (predicted_correct * 100)/messages_total
                        #create a list with score and f1
                        accuracy <- list(
                          "score" = score,
                          "f1" = f1,
                          "true_positives" = true_positives,
                          "true_negatives" = true_negatives,
                          "false_positives" = false_positives,
                          "false_negatives" = false_negatives
                        )
                        return(accuracy)
                    }
))

#0 - not spam
#1 - spam

model = naiveBayes()
p_list <- model$fit(training_X, training_y)

#example of prediction
prediction = model$predict("Go until jurong point, crazy.. Available only in bugis n great world la e buffet... Cine there got amore wat...", p_list[["first"]], p_list[["second"]])
print(prediction)

accuracy <- model$score(test_X, test_y)
sprintf("Score: %.2f", accuracy[["score"]])
sprintf("F1 score metric: %.2f", accuracy[["f1"]])
#print(model$incorrect)
```

## Measure effectiveness of your classifier

-   Note that accuracy is not always a good metric for your classifier.
    Look at precision and recall curves, F1 score metric.
-   Visualize them.
-   Show failure cases.
```{r}
all_data <- nrow(model$df_test)
incorrect_data <- nrow(model$incorrect)
correct_data <- all_data - incorrect_data

data_df <- data.frame(
  Value = c(correct_data, incorrect_data),
  Label = c("Correctly predicted", "Incorrectly predicted")
)
barplot(data_df$Value, names.arg = data_df$Label, col = c("green", "red"),
        main = "Prediction accuracy",
        xlab = "", ylab = "Count")

```
```{r}
data <- data.frame(
  Actual = c("Positive", "Negative", "Positive", "Negative"),
  Predicted = c("Positive", "Negative", "Negative", "Positive"),
  Value = c(accuracy[["true_positives"]], accuracy[["true_negatives"]], accuracy[["false_positives"]], accuracy[["false_negatives"]])
)

# Create a heatmap of the confusion matrix
ggplot(data, aes(x = Actual, y = Predicted, fill = Value)) +
  geom_tile() +
  geom_text(aes(label = Value), vjust = 1) +
  scale_fill_gradient(low = "white", high = "green") +
  labs(title = "Confusion Matrix", x = "Actual", y = "Predicted") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1))
```
```{r}
num_rows <- nrow(tidy_text_test)

all_data_df <- tidy_text_test
half_data_df <- tidy_text_test[1:(num_rows / 2), ]
quart_data_df <- tidy_text_test[1:(num_rows / 4), ]
one_e_data_df <- tidy_text_test[1:(num_rows / 8), ]
one_sixth_data_df <- tidy_text_test[1:(num_rows / 16), ]

res1 <- model$score(all_data_df$splitted, all_data_df$Category)
res2 <- model$score(half_data_df$splitted, half_data_df$Category)
res3 <- model$score(quart_data_df$splitted, quart_data_df$Category)
res4 <- model$score(one_e_data_df$splitted, one_e_data_df$Category)
res5 <- model$score(one_sixth_data_df$splitted, one_sixth_data_df$Category)

values <- c(res1[["score"]] / 100,
            res2[["score"]] / 100,
            res3[["score"]] / 100,
            res4[["score"]] / 100,
            res5[["score"]] / 100)

names <- c(toString(round((res1[["score"]] / 100), digits = 3)),
           toString(round((res2[["score"]] / 100), digits = 3)),
           toString(round((res3[["score"]] / 100), digits = 3)),
           toString(round((res4[["score"]] / 100), digits = 3)),
           toString(round((res5[["score"]] / 100), digits = 3)))

barplot(values, names.args = names, xlab = "Size of dataset", ylab = "Accuracies", col = ("green"))
axis(1, at = 1:length(names), labels = names)
```
```

## Conclusions

After our team have spent pretty huge amout of time writing and graphing
our naive Bayes model, we can highlight some thoughts:

-   First of all, our model predicts whether given message is spam or
    not spam with almost perfect accuracy (score \~ 0.96, F1 score \~
    0.85), but not good enough. For instance, on big dataflow, we will
    get quite big amount of incorrectly predicted messages.
-   Secondly, as barplot from above shows, size of dataset doesn't
    matter. Margin of error fluctuates between 0.1 and 0.001.
-   One more thing, we were convinced on our own experience, that
    usually simple things work perfectly on complicated problems.

